---
title: "Phyloseq Pre-Processing"
author: "Luna Eresta Jaya"
date: "2023-04-26"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Goals of our Pre-Processing File

We will work here in this document to remove any unnecessary ASVs or samples, such as:
1. Mitochondria ASVs
2. Chloroplast ASVs

## Load Packages and Functions

```{r load-packages-functions}
# Set wd
# setwd("/workdir/lej52/BIOMI6300_Project2_Amplicon_Analysis")

# Load packages
pacman::p_load(tidyverse, phyloseq, install = FALSE)

# Source functions
source("/local/workdir/lej52/BIOMI6300_Project2_Amplicon_Analysis/code/functions.R")
source("/local/workdir/lej52/BIOMI6300_Project2_Amplicon_Analysis/code/colors_and_shapes.R")
```


# Load the data

```{r load-data}
# Load in the raw_phyloseq data from DADA2_workflow.Rmd
load("/local/workdir/lej52/BIOMI6300_Project2_Amplicon_Analysis/data/raw_physeq.RData")

# Take a look at the S4 object!
raw_physeq

# Tell us about you phyloseq!
str(raw_physeq)
typeof(raw_physeq)

View(raw_physeq@otu_table)
View(raw_physeq@tax_table)
View(data.frame(sample_data(raw_physeq)))
```


## Remove unnecessary taxa!

```{r rm-ASVs}

noMitoChloro_physeq <-
  raw_physeq %>%
  # Remove the mitochondria ASVs
  subset_taxa(Family != "Mitochondria" | is.na(Family)) %>%
  # Remove the chloroplast ASVs
  subset_taxa(Order != "Chloroplast" | is.na(Order)) %>%
  # Now we will remove any ASVs of count 0
  prune_taxa(taxa_sums(.) > 0, .) # . is the holding space of our phyloseq objects that make it through our pipes

noMitoChloro_physeq

# How many taxa have we removed so far?
num_ASVs_rm <- ntaxa(raw_physeq) - ntaxa(noMitoChloro_physeq)


# Proportion of original data maintained
prop_ASVs_rm <- ntaxa(noMitoChloro_physeq)/ntaxa(raw_physeq)

```

Methods:

Mitochondria, chloroplasts and mock communities were removed resulting in the loss of `r  1-prop_ASVs_rm`% of the data.

## Check pre-processing results

```{r}
# Look at the phyloseq object
noMitoChloro_physeq

# Calculate the sum of the seqs in each sample
seqSums_df <-
  noMitoChloro_physeq %>%
  otu_table() %>%
  # Take the sum of each column of the ASV matrix
  colSums() %>%
  data.frame() %>%
  rownames_to_column(var = "names") # %>%
  # make_MA_metadata()

View(seqSums_df)

View(data.frame(noMitoChloro_physeq@otu_table))

# Fix column name in 2nd column
colnames(seqSums_df)[2] <- "TotalSeqs"

# Take a look at the data frame
head(seqSums_df)

# Plot it
# histogram plot
seqSums_df %>%
  ggplot(aes(x = TotalSeqs)) +
  geom_histogram() +
  labs(title = "Histogram of Raw Sequence Counts per sample")

# density plot
seqSums_df %>%
  ggplot(aes(x = TotalSeqs)) +
  geom_density() +
  labs(title = "Density plot of of Raw Sequence Counts per sample")

# # Density plots
# seqSums_df %>%
#   ggplot(aes(x = TotalSeqs, fill = fraction)) +
#   geom_density() +
#   scale_fill_manual(values = fraction_colors)
```

# Remove sample with too few reads
```{r}
preprocessed_physeq <-
  noMitoChloro_physeq # %>%
#  subset_samples(., names != "20210615-MA-SCB2F")

min(sample_sums(preprocessed_physeq))
```

# Save preprocessed phyloseq!
```{r save-physeq}
save(preprocessed_physeq, file = "/local/workdir/lej52/BIOMI6300_Project2_Amplicon_Analysis/data/preprocessed_physeq.RData")
```

